---
title: "Deseq2 report of `r alternate_condition` vs `r reference_condition`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r loading_libraries, include = FALSE}
#loading libraries
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(pheatmap))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(glmpca))
suppressPackageStartupMessages(library(apeglm))
suppressPackageStartupMessages(library(sva))
suppressPackageStartupMessages(library(RUVSeq))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(vsn))
suppressPackageStartupMessages(library(writexl))
suppressPackageStartupMessages(library(rafalib))
suppressPackageStartupMessages(library(ggfortify))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(cowplot))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(fgsea))
suppressPackageStartupMessages(library(msigdbr))
```

### Parameter used for the comparison\n
Here are the parameters listed, that are necessary for the comparison. 

```{r params_used, message=TRUE}
message("comparison_key: ",comparison_key)
message("reference_condition: ",reference_condition)
message("alternate_condition: ",alternate_condition)
if(any(other_keys!=FALSE)){
  message("other_keys: ",other_keys)
}
```
`r reference_condition` and `r alternate_condition` are groups of `r comparison_key` are compared against each other.

```{r, results='asis'}
if(any(other_keys!=FALSE)){
  cat("The comparison also considers", other_keys )
}
```

```{r writing_used_params}
# setwd(outdir)
filename <- sprintf("used_parameters_%s_vs_%s.txt", alternate_condition, reference_condition)
con <- file(filename, "w")
write_yaml(params, file=filename)
close(con)
```

```{r modifying_params, include=FALSE}
compared <- sprintf("%s_vs_%s", alternate_condition, reference_condition)
if(any(other_keys!=FALSE)){
  other_keys <- other_keys %>%
    strsplit(split = ",") %>%
    unlist()
} 
if(any(genes_of_interest!=FALSE)){
  genes_of_interest <- genes_of_interest %>%
      strsplit(split = ",") %>%
      unlist()
  }
if(any(perform_batch_correction!=FALSE)){
  perform_batch_correction <- perform_batch_correction %>%
      strsplit(split = ",") %>%
      unlist()
}

contrast <- list()
contrast <- c(comparison_key, alternate_condition, reference_condition)
coef <- sprintf("%s_%s", comparison_key, compared)
```

### Which groups should be compared and with with conditions and batches
We read in a count matrix, which we will name `cts`, and the sample information table, which we will name `coldata`. \n 
We examine the count matrix and column data to see if they are consistent in terms of sample order."

```{r building_coldata_cts, include=FALSE, message=TRUE}
# --- forming count and coldata
cts <- read.csv(count_data, sep = "\t", row.names = "gene_id")
coldata <- read.csv(metadata, row.names = 1) %>%
    dplyr::select(any_of(c(comparison_key, other_keys))) %>%
    dplyr::arrange(.data[[comparison_key]])
coldata[[comparison_key]] <- as.factor(coldata[[comparison_key]]) # Reordering so that the entries in the metadata and the countdata are in the same order
if(any(other_keys!=FALSE)){
  for(i in 1:length(other_keys)){
    coldata[[other_keys[i]]] <- as.factor(coldata[[other_keys[i]]])
  }
}
coldata[[comparison_key]] <- relevel(coldata[[comparison_key]], ref = reference_condition)

# This gets rid of the gene name column
cts <- cts[, rownames(coldata)]
if(!all(rownames(coldata) == colnames(cts))){
  stop("Rownames of coldata don't align with colnames of cts. Check the samplenames and their order in metadata and count_data.")
}

# --- adding colors into the dataframe coldata
comp_col <- unique(coldata[[comparison_key]])
colors <- brewer.pal(length(unique(comp_col)), "Set2")
names(colors) <- comp_col
coldata <- coldata %>%
    mutate(color = colors[.data[[comparison_key]]])


# --- color adaping if specific colors should be used
# comp_col <- unique(coldata[[comparison_key]])
# comp_col <- factor(conditions_ordered)
# colors <- c("#999999ff","#408002ff","#0f80ffff","#074080ff")
# names(colors) <- comp_col
# coldata <- coldata %>%
#     mutate(color = colors[.data[[comparison_key]]])


```

```{r printing_counts_coldata, echo=TRUE, messages=TRUE}
head(cts,5)
head(coldata,20)
```

```{r , include=FALSE}
######################################################## functions ########################################################
```

```{r functions, include=FALSE}
###############
# outlier detection
###############
# Figuring out sample-level outliers using plotRLE
check_n_remove_outliers <- function(cts = cts, coldata = coldata) {
  RLE_before <- plotRLE(round(as.matrix(cts)), outline = FALSE, ylim = c(-4, 4),returnData=TRUE)
  rle_stddevs <- apply(RLE_before, FUN = sd, MARGIN = 2)
  outliers <- rle_stddevs[rle_stddevs > 1] %>%
    names()
  to_keep <- rle_stddevs[rle_stddevs <= 1] %>%
    names()
  if(detect_sample_outliers){
    coldata <- coldata[to_keep, ]
    cts <- cts[, to_keep]
  }
  return(list(RLE_before, coldata, cts, outliers))
}


###############
# DE genes table
###############
calculate_DEG <- function(res = res, name = name){
  if(list_DEG){
    DEGs_res <- subset(res, res$padj<alpha)
    DEGs_res <- as.data.frame(DEGs_res)
    DEGs_res <- DEGs_res %>%
        mutate(gene_name = c(rownames(DEGs_res))) %>%
        relocate(gene_name, .before=1)
    tsv_DEG <- sprintf("%s/DEG_list_%s.tsv", outdir, name)
    write.table(DEGs_res, tsv_DEG, sep="\t", row.names = FALSE)
  }
}

listing_genes <- function(res = res, name = name){
  if(list_all_genes){
    res_df <- as.data.frame(res) %>%
        mutate(gene_name = c(rownames(res))) %>%
        relocate(gene_name, .before=1)
    tsv_allGenes <- sprintf("%s/geneList_%s.tsv", outdir, name)
    write.table(res_df, tsv_allGenes, sep="\t", row.names = FALSE)
  }
  return(tsv_allGenes)
}

###############
# meadn sd plot
###############
mean_sd_plot <- function(obj = obj, used = used){
  plot_mean_sd <- meanSdPlot(assay(obj), plot=FALSE)
  plot <- plot_mean_sd$gg + ggtitle(perform_variance_stabilisation)
  print(plot)
}


###############
# PCA
###############
perform_pca <- function(pcaData = pcaData, x=x, y=y, title = title){
  percentVar <- round(100 * attr(pcaData, "percentVar"))
  if(x=="PC1"||y=="PC1"){
    x_lab <- paste0(x,": ",percentVar[1],"% variance")
    y_lab <- paste0(y,": ",percentVar[2],"% variance") 
  } else{
    x_lab <- paste0(x)
    y_lab <- paste0(y)
  }
  if(any(other_keys!=FALSE)){
      pcaData[[other_keys[1]]] <- dds[[other_keys[1]]] %>% as.factor()
      plot_pca <- ggplot(pcaData, aes(x=.data[[x]], y=.data[[y]], color=.data[[comparison_key]], shape=.data[[other_keys[1]]], label=rownames(coldata)), returnData=TRUE) +
        geom_point(size=8) +
        geom_text(size=4,colour="black") +
        scale_color_manual(breaks = coldata[[comparison_key]], values = coldata$color) +
        labs(x = x_lab, y = y_lab, title = paste0("PCA - DESeq2 - ",title)) +
        theme_bw()+ 
        theme(axis.title.x = element_text(size =15),
          axis.text.x = element_text(size =12, angle=90),
          axis.text.y = element_text(size =12),
          axis.title.y = element_text(size = 15, angle = 90),
          plot.title=element_text(size = 16, face="bold"),
          legend.title = element_text(size = 14), 
          legend.text = element_text(size = 12))
    } else{
      plot_pca <- ggplot(pcaData, aes(x=.data[[x]], y=.data[[y]], color=.data[[comparison_key]], label=rownames(coldata)), returnData=TRUE) +
        geom_point(size=8) +
        geom_text(size=4,colour="black") +
        scale_color_manual(breaks = coldata[[comparison_key]], values = coldata$color) +
        labs(x = x_lab, y = y_lab, title = paste0("PCA - DESeq2 - ",title))+
        theme_bw()+ 
        theme(axis.title.x = element_text(size =15),
          axis.text.x = element_text(size =12, angle=90),
          axis.text.y = element_text(size =12),
          axis.title.y = element_text(size = 15, angle = 90),
          plot.title=element_text(size = 16, face="bold"),
          legend.title = element_text(size = 14), 
          legend.text = element_text(size = 12))
    }
  return(plot_pca)
}


###############
# counts
###############

plotting_counts <- function(dds=dds, genes_of_interest = genes_of_interest){
  if(any(counts != FALSE)){
    genes_to_plot <- list()
    if(any(genes_of_interest==FALSE)){
      genes_to_plot <- rownames(resOrdered[1:3,])
    } else{
      genes_to_plot <- genes_of_interest
    }
    genes_df <- list()
    genes_length <- length(genes_to_plot)

    if(any(counts == "barplot")){
      for(i in 1:genes_length){
        genes_df[[i]] <- plotCounts(dds, gene=genes_to_plot[i], intgroup=comparison_key, returnData = TRUE)
        genes_data <- genes_df[[i]]
        mean_sd_groups <- genes_data %>%
            group_by(.data[[comparison_key]]) %>%
            summarize(mean = mean(count), sd = sd(count))
        par(mfrow = c(3, 1), mar = c(3,5,3,1))
        plot_counts <- ggplot(data=mean_sd_groups, aes(x=.data[[comparison_key]], y=mean, fill = .data[[comparison_key]])) +
          geom_col(data=mean_sd_groups, aes(x=.data[[comparison_key]], y=mean)) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
          labs(x = comparison_key, y="normalized count", title=genes_to_plot[i]) +
          geom_point(data = genes_data, aes(x = .data[[comparison_key]], y=count, fill = .data[[comparison_key]])) +
          geom_errorbar(data=mean_sd_groups, aes(x=.data[[comparison_key]], ymin=mean-sd, ymax=mean+sd), width=0.4, alpha=0.9, size=0.5) +
          scale_fill_manual(breaks = coldata[[comparison_key]], values = coldata$color) +
          theme_bw() +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
          theme(axis.title.x = element_text(size =15),
          axis.text.x = element_text(size =12, angle=90),
          axis.text.y = element_text(size =12),
          axis.title.y = element_text(size = 15, angle = 90),
          plot.title=element_text(size = 16, face="bold"),
          legend.title = element_text(size = 14), 
          legend.text = element_text(size = 12)) 
        print(plot_counts, ncol=2)
      }
    }
    if(any(counts == "boxplot")){
      for(i in 1:genes_length){
        genes_df[[i]] <- plotCounts(dds, gene=genes_to_plot[i], intgroup=comparison_key, returnData = TRUE)
        genes_data <- genes_df[[i]]
        par(mfrow = c(3, 1), mar = c(3,5,3,1))
        plot_box <- ggplot(data=genes_data, aes(x=.data[[comparison_key]], y=count, ymin=0, fill=.data[[comparison_key]])) +
          geom_boxplot(outlier.shape = NA) +
          labs(x = comparison_key, y="normalized count", title=genes_to_plot[i]) +
          geom_point(data = genes_data, aes(x = .data[[comparison_key]], y=count, fill = .data[[comparison_key]])) +
          scale_fill_manual(breaks = coldata[[comparison_key]], values = coldata$color) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
          theme_bw()+
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
          theme(axis.title.x = element_text(size =15),
          axis.text.x = element_text(size =12, angle=90),
          axis.text.y = element_text(size =12),
          axis.title.y = element_text(size = 15, angle = 90),
          plot.title=element_text(size = 16, face="bold"),
          legend.title = element_text(size = 14), 
          legend.text = element_text(size = 12))
        print(plot_box)
      }
    }
  }
}
```


```{r , include=FALSE}
######################################################## caculations ########################################################
```


```{r outlier_detection, include=FALSE}

###############
# outlier detection
###############

# --- detection and removing of outliers
if (detect_sample_outliers) {
    result <- check_n_remove_outliers(cts, coldata) # using the created function check_n_remove_outliers
    RLE_before <- result[[1]]
    cts <- result[[3]]
    outliers <- result[[4]]

    # Plot RLE distribution using ggplot2, before outlier removal    
    df_rle_scores_before <- RLE_before %>%
        data.frame() %>%
        tidyr::pivot_longer(cols = where(is.numeric)) %>%
        mutate(condition = coldata[.data$name, comparison_key], color = coldata[.data$name, "color"]) %>%
        arrange(coldata[.data$name, comparison_key])
    df_rle_scores_before$alpha <- ifelse(df_rle_scores_before$name %in% outliers,0.2, 1)
    df_rle_scores_before$name <- factor(df_rle_scores_before$name, levels = rownames(coldata))
    plot_before <- ggplot(df_rle_scores_before, aes(x = name, y = value, fill = condition, alpha = alpha)) +
        geom_boxplot(outlier.shape = NA) +
        scale_alpha(guide = 'none') +
        scale_fill_manual(breaks = coldata[[comparison_key]], values = coldata$color) +
        ylim(c(-3,3)) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
        labs(x="sample", y="relative log expression", title = "RLE scores") + 
        theme(axis.title.x = element_text(size =15),
          axis.text.x = element_text(size =12, angle=90),
          axis.text.y = element_text(size =12),
          axis.title.y = element_text(size = 15, angle = 90),
          plot.title=element_text(size = 16, face="bold"),
          legend.title = element_text(size = 14), 
          legend.text = element_text(size = 12))
    
    coldata <- result[[2]]
    coldata <- coldata[order(coldata[[comparison_key]]), ]
    cts <- cts[, rownames(coldata)]
    if(!all(rownames(coldata) == colnames(cts))){
      stop("Rownames of coldata don't align with colnames of cts. Check the samplenames and their order in metadata and count_data.")
    }
    
}
```

```{r building_dds_object}
###############
# dds object
###############

if(any(other_keys!=FALSE)){
  design <- paste0("~ ", paste(c(other_keys, comparison_key), collapse = " + ")) %>% formula()
} else{
  design <- paste0("~ ", comparison_key) %>% formula()
}
dds <- DESeqDataSetFromMatrix(countData = round(cts), colData = coldata, design = design)

# --- Pre-filtering
minSamples <- min(table(coldata[[comparison_key]]))
keep <- rowSums(counts(dds) >= minCounts) >= minSamples
dds <- dds[keep, ]

dds <- DESeq(dds)

res <- results(dds, alpha=alpha, lfcThreshold=min_logfc_for_stat_test, altHypothesis="greaterAbs", contrast=contrast)
resOrdered <- res[order(res$padj),]

deg_results <- calculate_DEG(res, compared)

```

```{r DEG_xlsx}

###############
# DEG table
###############
if(list_DEG){
  deg_results <- calculate_DEG(res, compared)
}

if(list_all_genes){
  all_genes_results <- listing_genes(res, compared)
  tsv_allGenes <- all_genes_results[[1]]
}


```


```{r mean_sd, include=FALSE}

###############
# mean sd
###############
# --- stabilization 
if(perform_variance_stabilisation == FALSE){
  mean_vs_sd_after_stabilisation <- FALSE
  obj <- dds
  used <- "no stabilisation"
} else{
# vst
  if(perform_variance_stabilisation == "vst"){
    obj <- vst(dds, blind=do_blind_stabilization)
    used <- perform_variance_stabilisation
  }

# rlog
  if(perform_variance_stabilisation == "rlog"){
    obj <- rlog(dds, blind=do_blind_stabilization)
    used <- perform_variance_stabilisation
  }

# norm
  if(perform_variance_stabilisation == "norm"){
    obj <- normTransform(dds)
    used <- perform_variance_stabilisation
  }
}
```


```{r pca, include=FALSE}

###############
# PCA
###############

if(PCA){
  if(any(other_keys!=FALSE)){
    intgroup <- c(comparison_key, other_keys)
  } else{
    intgroup <- comparison_key
  }

  if(perform_variance_stabilisation == "vst"){
    pcaData <- plotPCA(obj, intgroup=intgroup,returnData=TRUE)
    title <- "vst"
  }

  if(perform_variance_stabilisation == "rlog"){
    pcaData <- plotPCA(obj, intgroup=intgroup,returnData=TRUE)
    title <- "rlog"
  } 
  
  if(perform_variance_stabilisation == "norm"){
    pcaData <- plotPCA(obj, intgroup=intgroup,returnData=TRUE)
    title <- "norm"
  } 
  
  if(perform_variance_stabilisation == FALSE){
    se <- SummarizedExperiment(log2(counts(dds, normalized=TRUE)),colData=colData(dds))
    pcaData <- plotPCA(DESeqTransform(se), intgroup=intgroup,returnData=TRUE)
    title <- "dds"
  }
  }
```


```{r lfc_shrink_MA_plot}

###############
# Shrink log fold change
###############

if(shrink_LFC_using == "apeglm"){
  resLFC <- lfcShrink(dds, coef = coef,type="apeglm", lfcThreshold=min_logfc_for_stat_test)
  res_shrink <- resLFC
}
if(shrink_LFC_using == "ashr"){
  resAsh <- lfcShrink(dds, coef = coef, type="ashr", lfcThreshold=min_logfc_for_stat_test)
  res_shrink <- resAsh
}
if(shrink_LFC_using == "normal"){
  resNorm <- lfcShrink(dds, coef = coef, type="normal", lfcThreshold=min_logfc_for_stat_test)
  res_shrink <- resNorm
}
```


```{r batch_correction, include=FALSE}
###############
# batch correction - RUV
###############

if(any(perform_batch_correction == "RUV")){
  set <- newSeqExpressionSet(counts(dds))
  set <- betweenLaneNormalization(set, which = "upper")
  not_sig <- rownames(res)[which(res$pvalue > RUV_threshold_not_sig)]
  empirical <- rownames(set)[rownames(set) %in% not_sig]
  set <- RUVg(set, empirical, k = 2)
  ddsruv <- dds
  ddsruv$W1 <- set$W_1
  ddsruv$W2 <- set$W_2
  coldata_ruv <- coldata
  coldata_ruv$W1 <- colData(ddsruv)$W1
  coldata_ruv$W2 <- colData(ddsruv)$W2
  if(any(other_keys!=FALSE)){
    design_ruv <- paste0("~ ", paste(c( other_keys, comparison_key, "W1 + W2"), collapse = " + ")) %>% formula()
  } else{
    design_ruv <- paste0("~ ", paste(c(comparison_key, "W1 + W2"), collapse = " + ")) %>% formula()
  }
  ddsruv <- DESeqDataSetFromMatrix(countData = round(cts), colData = coldata_ruv, design = design_ruv)
  ddsruv <- DESeq(ddsruv)
  resruv <- results(ddsruv, alpha=alpha, contrast=contrast)
  resruv
}

###############
# batch correction - SVA
###############

if(any(perform_batch_correction == "SVA")){
  dat <- counts(dds, normalized = TRUE)
  idx <- rowMeans(dat) > 1    #filtering out rows with a mean <= 1
  dat <- dat[idx, ]
  compare_col <- paste0("~ ", paste(comparison_key)) %>% formula()
  mm <- model.matrix(compare_col, colData(dds))
  mm0 <- model.matrix(~ 1, colData(dds))
  sva <- svaseq(dat, mod = mm, mod0 = mm0, n.sv = 2)
  ddssva <- dds
  ddssva$SV1 <- sva$sv[, 1]
  ddssva$SV2 <- sva$sv[, 2]
  coldata_sva <- coldata
  coldata_sva$SV1 <- colData(ddssva)$SV1
  coldata_sva$SV2 <- colData(ddssva)$SV2
  if(any(other_keys!=FALSE)){
    design_sva <- paste0("~ ", paste(c(other_keys, comparison_key, "SV1 + SV2"), collapse = " + ")) %>% formula()
  } else{
    design_sva <- paste0("~ ",paste(c(comparison_key, "SV1 + SV2"), collapse = " + ")) %>% formula()
  }
  ddssva <- DESeqDataSetFromMatrix(countData = round(cts), colData = coldata_sva, design = design_sva)
  ddssva <- DESeq(ddssva)
  ressva <- results(ddssva, alpha=alpha, contrast=contrast)
  ressva
}
```


```{r , include=FALSE}
######################################################## PLOTS ########################################################
```

## Differential expression analysis
With the count matrix, cts, and the sample information, coldata, we can construct a DESeqDataSet.
The standard differential expression analysis steps are wrapped into a single function, DESeq. 
The estimation steps performed by this function are described below, in the manual page for ?DESeq and in the Methods section of the DESeq2 publication (Love, Huber, and Anders 2014).

Here we perform pre-filtering to keep only rows that have a count of at least `r minCounts` for a minimal number of samples
in at least `r minSamples` groups. 

```{r, message = TRUE} 
message("design: ",design)
```

```{r, results='asis', eval=detect_sample_outliers}
cat("##  Filtering and exploratory data analysis\n
The boxplot of relative log expression (RLE) (log-ratio of read count to median read count across the samples).\n
Detected outliers are marked as transparent bars.")
```

```{r plot_outliers, fig.width=10, fig.height=6, dpi=300, eval=detect_sample_outliers}
  print(plot_before)
```


```{r, results='asis', eval=PCA}
cat("## Principal component plot of the samples\n
Related to the distance matrix is the PCA plot, which shows the samples in the 2D plane spanned 
by their first two principal components. This type of plot is useful for visualizing 
the overall effect of experimental covariates and batch effects.")
```


```{r plot_pca, fig.width=8, fig.height=6, dpi=300, eval=PCA}

pcaPlot <- perform_pca(pcaData, x = "PC1", y = "PC2", title)
print(pcaPlot)

if(any(perform_batch_correction == "RUV")){
  se_ruv <- SummarizedExperiment(log2(counts(ddsruv, normalized=TRUE)),colData=colData(ddsruv))
  pcaData_ruv <- plotPCA(DESeqTransform(se_ruv), intgroup=intgroup,returnData=TRUE)
  title <- "dds_ruv"
  pcaPlot_ruv <- perform_pca(pcaData_ruv, x = "PC1", y = "PC2", title="RUV")
  print(pcaPlot_ruv)
}

if(any(perform_batch_correction == "SVA")){
  se_sva <- SummarizedExperiment(log2(counts(ddssva, normalized=TRUE)),colData=colData(ddssva))
  pcaData_sva <- plotPCA(DESeqTransform(se_sva), intgroup=intgroup,returnData=TRUE)
  title <- "dds_sva"
  pcaPlot_sva <- perform_pca(pcaData_sva, x = "PC1", y = "PC2", title="SVA")
  print(pcaPlot_sva)
}
  
```

## Histogram of p values for all tests.
Showing the histogram of pvalue without adjustment and of adjusted pvalue using the in DESeq2 implemented Benjamini-Hochberg procedure.

```{r hist_pvalues, fig.width=8, fig.height=6, dpi=300}
par(mfrow=c(1,2), mar = c(5,4,4,2))
hist(res$pvalue, breaks=0:40/40, main=sprintf("%s \n - my analysis \n - non-adjusted pvalue",compared), xlim=c(0,1), cex.main=1, xlab="p-value")
hist(res$padj, breaks=0:40/40, main=sprintf("%s \n - my analysis \n - adjusted pvalue",compared), xlim=c(0,1), cex.main=1, xlab="adjusted p-value")

```

```{r plot_RUV, results='asis', fig.width=8, fig.height=6, dpi=300}
if(any(perform_batch_correction == "RUV")){
  cat("### RUV Histograms \n")
  par(mfrow = c(1, 2))
  hist(resruv$pvalue, breaks=0:40/40, main=sprintf("%s - vs -\n %s \n - non-adjusted pvalue - RUV",alternate_condition,reference_condition), cex.main=.8, xlab="p-value")
  hist(resruv$padj, breaks=0:40/40, main=sprintf("%s - vs -\n %s \n - adjusted pvalue - RUV",alternate_condition,reference_condition), cex.main=.8, xlab="adjusted p-value")
}
```

```{r plot_SVA, results='asis', fig.width=8, fig.height=6, dpi=300}
if(any(perform_batch_correction == "SVA")){
  cat("### SVA Histograms \n")
  par(mfrow = c(1, 2))
  hist(ressva$pvalue, breaks=0:40/40, main=sprintf("%s - vs -\n %s \n - non-adjusted pvalue - SVA",alternate_condition,reference_condition), cex.main=.9, xlab="p-value")
  hist(ressva$padj, breaks=0:40/40, main=sprintf("%s - vs -\n %s \n - adjusted pvalue - SVA",alternate_condition,reference_condition), cex.main=.9, xlab="adjusted p-value")
}
```

### DE genes 
A table with all DEGs is saved as ``r sprintf("DEG_list_%s.xlsx",compared)`` and all genes are saved in ``r sprintf("geneList_%s.xlsx",compared)``.
The amount of adjusted p-values less than `r alpha`
```{r DEG_table, messages=TRUE}
cat("without batch correction \n")
print(sprintf("DEGs with adjusted p-value < %g of %s: %d", alpha,compared,sum(res$padj < alpha, na.rm=TRUE)))
print(sprintf("DEGs with non-adjusted p-value < %g of %s: %d", alpha,compared,sum(res$pvalue < alpha, na.rm=TRUE)))

if(any(perform_batch_correction == "RUV")){
cat("RUV \n")
print(sprintf("DEGs with adjusted p-value < %g after batch correction using RUV of %s: %d", alpha,compared,sum(resruv$padj < alpha, na.rm=TRUE)))
print(sprintf("DEGs with non-adjusted p-value < %g after batch correction using RUV of %s: %d", alpha,compared,sum(resruv$pvalue < alpha, na.rm=TRUE)))
}

if(any(perform_batch_correction == "SVA")){
cat("SVA \n")
print(sprintf("DEGs with adjusted p-value < %g after batch correction using SVA of %s: %d", alpha,compared,sum(ressva$padj < alpha, na.rm=TRUE)))
print(sprintf("DEGs with non-adjusted p-value < %g after batch correction using SVA of %s: %d", alpha,compared,sum(ressva$pvalue < alpha, na.rm=TRUE)))
}
```


```{r, results='asis'}
if(shrink_LFC_using!=FALSE){
  cat("## LFC shrinkage\n
  Shrinkage of effect size (LFC estimates) is useful for visualization and ranking of genes. 
  To shrink the LFC, we pass the dds object to the function lfcShrink. 
  Below we specify to use the apeglm method for effect size shrinkage (Zhu, Ibrahim, and Love 2018), which improves on the previous estimator.
  In DESeq2, the function plotMA shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet. 
  Points will be colored blue if the adjusted p value is less than", alpha,".
  Points which fall out of the window are plotted as open triangles pointing either up or down.")
}

```

```{r lfc_shrink_MA_ploting, fig.width=10, fig.height=6, dpi=300}
par(mfrow=c(1,2), mar = c(5,4,4,2))

if(plot_MA_before_LFCshrink){
#the function plotMA shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet
  ylim <- c(-3,3)
# --- before shrinkage
  if(plot_MA_before_LFCshrink) {  
    MA_plot_before <- DESeq2::plotMA(res, ylim=ylim, main=paste0(compared, " - no shrinkage"), cex.main=.65)
    abline(h=c(-min_logfc_for_stat_test,min_logfc_for_stat_test), col="dodgerblue", lwd=2)
  }

# --- shrinkage
  if(shrink_LFC_using == "apeglm"){
    if(plot_MA_after_LFCshrink){
      MA_plot_after <- DESeq2::plotMA(resLFC, ylim=ylim, main=paste0(compared, " - apeglm"), cex.main=.65)
      abline(h=c(-min_logfc_for_stat_test,min_logfc_for_stat_test), col="dodgerblue", lwd=2)
    }
  }

  if(shrink_LFC_using == "ashr"){
    if(plot_MA_after_LFCshrink){
      MA_plot_after <- DESeq2::plotMA(resAsh, ylim=ylim, main=paste0(compared, " - ashr"), cex.main=.65)
      abline(h=c(-min_logfc_for_stat_test,min_logfc_for_stat_test), col="dodgerblue", lwd=2)
    }
  }

  if(shrink_LFC_using == "normal"){
    if(plot_MA_after_LFCshrink){
      MA_plot_after <- DESeq2::plotMA(resNorm, ylim=ylim, main=paste0(compared, " - normal"), cex.main=.65)
      abline(h=c(-min_logfc_for_stat_test,min_logfc_for_stat_test), col="dodgerblue", lwd=2)
    }
  }
}

```

```{r, results='asis', eval=counts}
if(any(counts != FALSE)){
cat("## Plot counts \n
It can also be useful to examine the counts of reads for a single gene across the groups. 
A simple function for making this plot is plotCounts, which normalizes counts by the estimated size factors 
(or normalization factors if these were used) and adds a pseudocount of 1/2 to allow for log scale plotting. 
The counts are grouped by the variables in intgroup, where more than one variable can be specified. 
Here we specify the gene which had the smallest p value from the results table created above. 
You can select the gene to plot by rowname or by numeric index.")
}
```


```{r plot_counts, fig.width=10, fig.height=6, dpi=300}
if(any(counts!=FALSE)){
  par(mfrow = c(3, 1), mar = c(3,5,3,1))
  plotting_counts(dds, genes_of_interest)
}
```

```{r, results='asis', eval=plot_sample_clustering}
cat("## Heatmap of the count matrix\n
To explore a count matrix, it is often instructive to look at it as a heatmap. ")
```


```{r plot_heatmap, eval=plot_sample_clustering, fig.width=8, fig.height=6, dpi=300}
ann_colors <- list()
ann_colors[[comparison_key]] <- colors
if(any(other_keys!=FALSE)){
  df <- as.data.frame(colData(obj)[c(comparison_key,other_keys)])
  for(i in 1:length(other_keys)){
    other_col <- unique(coldata[[other_keys[i]]])
    other_color <- colorRampPalette(brewer.pal(9, "Set1"))(length(unique(other_col)))
    if(length(other_col) < 3){
      other_color <- c("#03bfc4ff", "#f8766dff")
    }
    names(other_color) <- other_col
    ann_colors[[other_keys[i]]] <- other_color
  }
} else{
  df <- as.data.frame(colData(obj)[comparison_key])
}
topVarGenes <- order(-rowVars(assay(obj)))[0:200]
select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:200]

topRank <- match(rownames(resOrdered)[0:200],rownames(assay(obj)))

pheatmap(assay(obj)[topRank,],cluster_rows=FALSE, show_rownames=FALSE, cluster_cols=TRUE, annotation_col=df, annotation_colors = ann_colors, main=sprintf("Heatmap - %s \n - %s",used,compared))

```

```{r, results='asis', eval=plot_sample_clustering}
cat("## Heatmap of the sample-to-sample distances\n
Here we produce such a heatmap for various transformations of the data.
Another use of the transformed data is sample clustering. We apply the dist function to the transpose of the transformed count matrix to get sample-to-sample distances.")
```

```{r plot_heatmapsample_to_sample_distance, eval=plot_sample_clustering, fig.width=10, fig.height=6, dpi=300}
sampleDists <- dist(t(assay(obj)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(obj[[comparison_key]])
colnames(sampleDistMatrix) <- paste(colnames(obj))
pheatmap(sampleDistMatrix, clustering_distance_rows=sampleDists, clustering_distance_cols=sampleDists,main=sprintf("Euclidean distance - %s",used))
```

```{r gsea, eval=perform_gsea}
gene_ranks <- res_shrink %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "gene") %>%
  filter(!is.na(padj)) %>%
  arrange(desc(log2FoldChange)) %>%
  select(gene, log2FoldChange) %>%
  tibble::deframe()

pathways <- msigdbr(species = species) %>%
  filter(grepl(pattern = geneset_subset_pattern, .data[[geneset_subset_keys]], ignore.case = TRUE)) %>%
  base::split(x = .$ensembl_gene, f = .$gs_name)

fgseaRes <- fgsea(pathways = pathways,
	stats = gene_ranks,
	minSize  = geneset_minsize,
	maxSize  = geneset_maxsize)

rds_gsea <- sprintf("%s/fgsea_significant_genesets_%s_%s.rds", outdir, geneset_subset_keys, geneset_subset_pattern)
saveRDS(fgseaRes,file = rds_gsea)

fgsea_significant <- fgseaRes %>%
  filter(padj <= alpha)

# Save table of collapsed pathways 
collapsedPathways <- collapsePathways(fgseaRes = fgsea_significant, 
  pathways = pathways,
  stats = gene_ranks,
  pval.threshold = alpha)
mainPathways <- fgseaRes[pathway %in% collapsedPathways$mainPathways] %>%
  arrange(desc(NES)) %>%
	select(-where(is.list))

tsv_gsea <- sprintf("%s/fgsea_significant_genesets_%s_%s.tsv", outdir, geneset_subset_keys, geneset_subset_pattern)
write.table(mainPathways, tsv_gsea, sep="\t", row.names = FALSE)
```

# More detailed information
## Results
Results tables are generated using the function results, which extracts a results table with log2 fold changes, 
p values and adjusted p values. With no additional arguments to results, the log2 fold change and Wald test 
p value will be for the last variable in the design formula, and if this is a factor, the comparison will be 
the last level of this variable over the reference level (see previous note on factor levels). 
However, the order of the variables of the design do not matter so long as the user specifies the comparison 
to build a results table for, using the name or contrast arguments of results.

```{r min_logfc_for_stat_test}
message("min_logfc_for_stat_test: ", min_logfc_for_stat_test)
```

More information about the result table. In summary and which variables and tests were used can be found by calling the function mcols on the results object.
```{r results_description, echo=TRUE, messages=TRUE}
summary(res)
mcols(res)$description
```


```{r, results='asis', eval=dispersion}
cat("## Dispersion plot\n
Plotting the dispersion estimates is a useful diagnostic. The dispersion plot below is typical, 
with the final estimates shrunk from the gene-wise estimates towards the fitted estimates. 
Some gene-wise estimates are flagged as outliers and not shrunk towards the fitted value, 
(this outlier detection is described in the manual page for estimateDispersionsMAP). 
The amount of shrinkage can be more or less than seen here, depending on the sample size, 
the number of coefficients, the row mean and the variability of the gene-wise estimates.")
```

```{r dispersion_plot, eval=dispersion}
  plotDispEsts(dds, main=compared)
```

```{r, results='asis', eval=mean_vs_sd}
cat("## Data transformations and visualization\n
These transformation functions return an object of class DESeqTransform which is a subclass of RangedSummarizedExperiment. 
For ~20 samples, running on a newly created DESeqDataSet, rlog may take 30 seconds, while vst takes less than 1 second. 
The running times are shorter when using blind=FALSE and if the function DESeq has already been run, 
because then it is not necessary to re-estimate the dispersion values. \n

The figure below plots the standard deviation of the transformed data, across samples, against the mean, 
using the shifted logarithm transformation, the regularized log transformation and the variance stabilizing transformation. 
The shifted logarithm has elevated standard deviation in the lower count range, and the regularized log to a lesser extent, 
while for the variance stabilized data the standard deviation is roughly constant along the whole dynamic range. \n

Note that the vertical axis in such plots is the square root of the variance over all samples, 
so including the variance due to the experimental conditions. While a flat curve of the square root of variance 
over the mean may seem like the goal of such transformations, this may be unreasonable in the case of datasets with 
many true differences due to the experimental conditions.")
```


```{r plot_mean_sd, echo=TRUE, eval=mean_vs_sd}
  mean_sd_plot(obj, used)
```

```{r, results='asis', eval=volcano}
cat("## Volcano plot\n")
```

```{r volcano plots, eval=volcano}
plot(res$log2FoldChange, -log10(res$padj), panel.first=grid(), main="Volcano plot", xlab="Effect size: log2(fold-change)", ylab="-log10(adjusted p-value)", xlim=c(-4,4), pch=20, cex=0.6)
```

```{r, results='asis'}
cat("## Independent Filtering and multiple testing\n
The goal of independent filtering is to filter out those tests from the procedure that have no, or little chance of showing significant evidence, 
without even looking at their test statistic. Typically, this results in increased detection power at the same experiment-wide type I error. 
Here, we measure experiment-wide type I error in terms of the false discovery rate.

The results function of the DESeq2 package performs independent filtering by default using the mean of normalized counts as a filter statistic. 
A threshold on the filter statistic is found which optimizes the number of adjusted p values lower than a significance level alpha 
(we use the standard variable name for significance level, though it is unrelated to the dispersion parameter). 
The adjusted p values for the genes which do not pass the filter threshold are set to NA.

The default independent filtering is performed using the filtered_p function of the genefilter package, 
and all of the arguments of filtered_p can be passed to the results function. The filter threshold value and 
the number of rejections at each quantile of the filter statistic are available as metadata of the object returned by results.

For example, we can visualize the optimization by plotting the filterNumRej attribute of the results object. 
The results function maximizes the number of rejections (adjusted p value less than a significance level), 
over the quantiles of a filter statistic (the mean of normalized counts). The threshold chosen (vertical line) 
is the lowest quantile of the filter for which the number of rejections is within 1 residual standard deviation 
to the peak of a curve fit to the number of rejections over the filter quantiles:")
```

```{r independent_filtering}
plot(metadata(res)$filterNumRej, 
     type="b", ylab="number of rejections",
     xlab="quantiles of filter")
lines(metadata(res)$lo.fit, col="red")
abline(v=metadata(res)$filterTheta)
resNoFilt <- results(dds, alpha=alpha, lfcThreshold=min_logfc_for_stat_test, , altHypothesis="greaterAbs", contrast=contrast, independentFiltering=FALSE)
addmargins(table(filtering=(res$padj < alpha),
                 noFiltering=(resNoFilt$padj < alpha)))
```

```{r, results='asis'}
cat("A simple filtering criterion readily available in the results object is the mean of normalized counts irrespective of biological condition, 
and so this is the criterion which is used automatically by the results function to perform independent filtering. Genes with very low counts are not likely 
to see significant differences typically due to high dispersion. For example, we can plot the -log10 p values from all genes over the normalized mean counts:")
```

```{r plot_normCounts_pvalue, eval=normCounts_pvalue}
plot(res$baseMean+1, -log10(res$pvalue),
     log="x", xlab="mean of normalized counts",
     ylab=expression(-log[10](pvalue)),
     ylim=c(0,30),
     cex=.4)
```

```{r, results='asis', eval=count_outlier_detection}
cat("## Count outlier detection\n
DESeq2 relies on the negative binomial distribution to make estimates and perform statistical inference on differences. 
While the negative binomial is versatile in having a mean and dispersion parameter, extreme counts in individual samples might not fit well to the negative binomial. 
For this reason, we perform automatic detection of count outliers. We use Cook's distance, which is a measure of how much the fitted coefficients would change if an individual sample were removed (Cook 1977). 
For more on the implementation of Cook's distance see the manual page for the results function. 
Below we plot the maximum value of Cook's distance for each row over the rank of the test statistic to justify its use as a filtering criterion.")
```

```{r plot_count_outlier_detection, eval=count_outlier_detection}
maxCooks <- apply(assays(dds)[["cooks"]],1,max)
idx <- !is.na(res$stat)
plot(rank(res$stat[idx]), maxCooks[idx], xlab="rank of Wald statistic", 
     ylab="maximum Cook's distance per gene",
     ylim=c(0,5), cex=.4)
```

```{r, results='asis', eval=perform_batch_correction}
cat("## Batch correction \n")
if(!any(perform_batch_correction == "RUV") & !any(perform_batch_correction == "SVA")){
  cat(" No batch correction tool (RUV or SVA) was selected.")
}
```

```{r, results='asis'}
if(any(perform_batch_correction == "RUV")){
  cat("## RUV \n
  Plots of principal components (PC) reveal a clear need for betwen-sample normalization.
  We can also use the RUV method in the RUVSeq package to detect the hidden batch effects.
  We can use the RUVg function to estimate factors of unwanted variation, analogous to SVA's surrogate variables. 
  A difference compared to the SVA procedure, is that we first would run DESeq and results to obtain 
  the p-values for the analysis without knowing about the batches. 
  Supposing that we have this results table `res`, we then pull out a set of empirical control genes 
  by looking at the genes that do not have a small p-value.")
}
```

```{r, results='asis'}
if(any(perform_batch_correction == "RUV")){
  cat("We can plot the factors estimated by RUV:")
}
```

```{r plots_RUV}
if(any(perform_batch_correction == "RUV")){

  pData(set)
  par(mfrow = c(2, 1), mar = c(3,5,3,1))
  for (i in 1:2) {
    stripchart(pData(set)[, i] ~ dds[[comparison_key]], vertical = TRUE, main = paste0("W", i)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 
    abline(h = 0)
   }

  alpha_res <- print(sprintf("DEGs with p-value < %g of %s: %d", alpha,compared,sum(resruv$padj < alpha, na.rm=TRUE)))
  resNoFilt_ruv <- results(ddsruv, alpha=alpha, independentFiltering=FALSE)
  addmargins(table(filtering=(resruv$padj < alpha),
                 noFiltering=(resNoFilt_ruv$padj < alpha)))
  
  name <- paste0("ruv_",compared)
  
  deg_results_ruv <- calculate_DEG(res = resruv, name)
  deg_names_ruv <- deg_results_ruv[[1]]
  listing_genes(resruv, name)
}
```

```{r, results='asis', eval=perform_batch_correction}
if(any(perform_batch_correction == "SVA")){
  cat("## SVA \n
  Below we obtain a matrix of normalized counts for which the average count across samples is larger than 1. 
  As we described above, we are trying to recover any hidden batch effects. 
  So we use a full model matrix and a reduced, or null, model matrix with only an intercept term. 
  Finally we specify that we want to estimate 2 surrogate variables. 
  Finally, in order to use SVA to remove any effect on the counts from our surrogate variables, 
  we simply add these two surrogate variables as columns to the DESeqDataSet and then add them to the design:")
}
```

```{r plots_SVA}
if(any(perform_batch_correction == "SVA")){
  
  par(mfrow = c(2, 1), mar = c(3,5,3,1))
  for (i in 1:2) {
    stripchart(sva$sv[, i] ~ dds[[comparison_key]], vertical = TRUE, main = paste0("SV", i))
    abline(h = 0)
   }

  alpha_res <- print(sprintf("DEGs with p-value < %g of %s: %d", alpha,compared,sum(ressva$padj < alpha, na.rm=TRUE)))
  resNoFilt_sva <- results(ddssva, alpha=alpha, independentFiltering=FALSE)
  addmargins(table(filtering=(ressva$padj < alpha),
                 noFiltering=(resNoFilt_sva$padj < alpha)))
  name <- paste0("sva_",compared)
  
  deg_results_sva <- calculate_DEG(res = ressva, name)
  deg_names_sva <- deg_results_sva[[1]]
  listing_genes(ressva, name)
}
```
